---
title: "Homework 3"
date: "2026-05-25"
author: "Luke Simon"
format: pdf:
  toc: true
execute:
  message: false
  warning: false
---

```{r}
#| label: setup

# read in packages
library(tidyverse)
library(here)
library(janitor)
library(readxl)
library(patchwork)  

# read in data and create new object
kelp <- read_csv(
  here("data", "temp-kelp.csv"))
```


# Problem 1. Giant kelp fronds

## a.

There are two appropriate tests for testing the strength of the relationship between temperature and giant kelp frond elongation rate (correlation). These two are the Pearson's correlation, which assumes normally distributed continuous data, and Spearman's rank correlation, which does not require normality but does require continuous data. Pearsons'scorrelation is the correct test to choose when there are no outliers in the data and the distribution is clearly normal, in all other cases Spearman's is more appropriate.

## b.

```{r}
#| label: temp-elong-correlation-plot

# base layer: ggplot, data from kelp data frame
ggplot(data = kelp,
       mapping = aes(
         x = temp_c, # x-axis: temperature
         y = kelp_elong, # y-axis: elongation rate
         color = temp_c)) + # color gradient by temperature
  # first layer: scatter plot
  geom_point() +
  # labeling axes and adding title
  labs(x = expression(Temperature~(degree*C)),
       y = expression("Giant kelp elongation rate (cm day"^-1~")"),
       title = "As temperature increases, elongation rate decreases") +
  # changing theme
  theme_bw() +
  # removing legend
  theme(legend.position = "none")
```

## c.

### Assumption check

**Relationship appears linear, likely meaning Pearson's is the accurate test to run. Going to check normality of each variable, as each is continuous, to confirm that Pearson's is the correct test.**

```{r}
#| label: QQ-plots

# creating object for temperature QQ plot
tempQQ <-
# baselayer: ggplot
ggplot(data = kelp, # using kelp data frame
       aes(sample = temp_c)) + # y-axis for QQ plot
  # first layer: QQ reference line
  geom_qq_line(color = "cornflowerblue") + # displaying line in blue
  # second layer: QQ plot 
  geom_qq() +
  # labeling title
  labs(title = "Temperature QQ")+
  # changing theme
  theme_bw() 

# creating obhect for elongation QQ plot
elongQQ <-
# baselayer: ggplot
ggplot(data = kelp, # using kelp data frame
       aes(sample = kelp_elong)) + # y-axis for QQ plot
  # first layer: QQ reference line
  geom_qq_line(color = "red") + # displaying line in red
  # second layer: QQ plot 
  geom_qq() +
  # labeling title 
  labs(title = "Kelp elongation QQ") +
  # changing theme
  theme_bw()

# using patchwork package to display plots side-by-side
tempQQ | elongQQ
```

To figure out which test was accurate to run in this scenario I first checked the linearity of the initial plot, there appears to be a negative linear relationship between temperature and giant kelp elongation rate. Then, to check the normality (I am assuming this data is independent here and we have established that they both are continuous) I created QQ plots of each variable. The QQ plots suggest both variables (and therefore all the data) are normally distributed (centering around the reference line with only slight deviations at the tails) and therefore suggests that a Pearson's correlation test is the appropriate test to perform.

### Running the test

```{r}

# running correlation test
cor.test(kelp$temp_c, kelp$kelp_elong,
         method = "pearson") # pearson's test
```

## d.

To evaluate the strength of the relationship between temperature and giant kelp frond elongation rate, I uses a Pearson's correlation test. I found a strong negative relationship between temperature and elongation rate (Pearson's r = -0.69, t(30) = -5.2, p < 0.001, $\alpha$ =0.05)