---
title: "Homework 3"
date: "2026-05-25"
author: "Luke Simon"
format: pdf:
  toc: true
execute:
  message: false
  warning: false
---

```{r}
#| label: setup

# read in packages
library(tidyverse)
library(here)
library(janitor)
library(readxl)

# read in data and create new object
kelp <- read_csv(
  here("data", "temp-kelp.csv"))
```


# Problem 1. Giant kelp fronds

## a.

There are two appropriate tests for testing the strength of the relationship between temperature and giant kelp frond elongation rate (correlation). These two are the Pearson's correlation, which assumes normally distributed continuous data, and Spearman's rank correlation, which does not require normality but does require continuous data. Pearsons'scorrelation is the correct test to choose when there are no outliers in the data and the distribution is clearly normal, in all other cases Spearman's is more appropriate.

## b.

```{r}
#| label: temp-elong-correlation-plot

# base layer: ggplot, data from kelp data frame
ggplot(data = kelp,
       mapping = aes(
         x = temp_c, # x-axis: temperature
         y = kelp_elong, # y-axis: elongation rate
         color = temp_c)) + # color gradient by temperature
  # first layer: scatter plot
  geom_point() +
  # labeling axes and adding title
  labs(x = expression(Temperature~(degree*C)),
       y = expression("Giant kelp elongation rate (cm day"^-1~")"),
       title = "As temperature increases, elongation rate decreases") +
  # changing theme
  theme_bw() +
  # removing legend
  theme(legend.position = "none")
```

